---
title: "rancovr quickstart"
author: "Massimo Cavallaro"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# devtools::load_all()
```



The package `rancovr` provides a family of functions for detecting anomalies with the RAndom Neighbourhood COVeRing.
This method is particularly useful in epidemiological studies where one needs to evaluate whether a set of infection events occurring in a certain space and time is of concern and deviates from a null model prediction. In other words
 `rancovr` can help researchers compare observed data against null model predictions and determine whether the events are atypical or deviate significantly from what would be expected under the null model.
 
This vignette only provides a brief overview of the package. For a more comprehensive understanding of the methodology and its applications, we refer to references [[1]](#1) and [[2]].
The package includes two primary functions,  `CreateCylinders` and `warning.score`.
The `CreateCylinders` function is used to generate cylindrical shapes that cover the spatio-temporal events of interest. Once the cylindrical shapes have been generated, the `warning.score` function can be used to aggregate the information from all cylinders and calculate a continuous score between zero and one for each event. This score provides a measure of the event's deviation from the expected behavior under a given null model.


## Create dataset

The RAndom Neighbourhood COVeRing is designed to compare counts of discrete events with a baseline. In `rancovr` the baseline can be defined using a spatio-temporal grid, where the spatial component is expanded into rows using the `expand.grid` function (built-in in base R) and the columns represent time steps.

```{r}
x <- seq(0, 10, length.out = 20)
y <- seq(0, 10, length.out = 20)
tt = seq(1, 20)
d1 <- expand.grid(x = x, y = y)
d1$names = paste('loc', as.character(1:20),  sep = '-')
baseline.matrix = sapply(tt, function( . ){
  exp(- (d1$x - 5)^2 - (d1$y - 5)^2 )  *  cos( 0.2 * . )^2} )
rownames(baseline.matrix) = d1$names
colnames(baseline.matrix) = as.character(tt)
baseline.matrix[1:5, 1:5]
```

The values of `baseline.matrix` at row `i` and column `j` 
represent the expected number of events at location `i` during time step `j`.
To visualize the baseline at a specific time, the built-in R function `image` can be used to plot the matrix. In the resulting plot, darker colors indicate higher expected event counts.


```{r, fig.show='hold'}
image(x, y, matrix(baseline.matrix[,5], ncol=20), main='t=5')
image(x, y, matrix(baseline.matrix[,15], ncol=20), main='t=15')
```


To generate observations from this baseline, we   use the built-in R function `rpois`. Since these observations are generated from the baseline, they should not raise any alarms.


```{r}
observation.matrix = matrix(
  rpois(length(baseline.matrix), baseline.matrix),
  ncol = ncol(baseline.matrix))

rownames(observation.matrix) = d1$names
colnames(observation.matrix) = as.character(tt)
```


We can use the `image` function to illustrate the distribution of events generated from the baseline and compare it with the original baseline.
```{r, fig.show='hold'}
image(x, y, matrix(observation.matrix[,5], ncol=20), main='t=5')
image(x, y, matrix(observation.matrix[,15], ncol=20), main='t=15')
```

The aggregated observations and baseline are similar, within statistical fluctuations.

```{r, fig.width=5, fig.height=4}
plot(colSums(observation.matrix), ylab='', xlab='t')
lines(colSums(baseline.matrix))
legend('bottomright', c('observation', 'baseline'), pch=c(1, NA), lty=c(NA, 1))
```


We will now introduce a set of discrete events at selected locations and times that do not conform to the known baseline. This set of events will be referred to as a "novel cluster", and it is important that the algorithm is capable of detecting these points.

```{r}
observation.matrix[30:35, 10:15] = observation.matrix[30:35, 10:15] + rpois(36, 1)
```



```{r, fig.show='hold'}
t = 5
image(x, y, matrix(observation.matrix[,t], ncol=20), main = paste0('t=', as.character(t)))
t = 15
image(x, y, matrix(observation.matrix[,t], ncol=20), main = paste0('t=', as.character(t)))
```

```{r, fig.width=5, fig.height=4}
plot(colSums(observation.matrix), ylab='', xlab='t')
lines(colSums(baseline.matrix))
legend('bottomright', c('observation', 'baseline'), pch=c(1, NA), lty=c(NA, 1))
```

## Retrospective usage

We now utilise `rancovr` to identify the cluster of atypical events. 
Retrospectively, we generate all possible cylinders up until a specified date `TT` using the function `CreateCylinders` with the `only.last=FALSE` option.

```
cylinders = CreateCylinders(observation.matrix,
                            baseline.matrix, range(tt), n.cylinders = 1000,
                            coord.df = d1, size_factor = 4, GT = 5)
head(cylinders)
```



Each cylinder encompasses a number of observed events. The expected number of events in each cylinder is determined by the Poisson statistics and can be plotted against the actual number of observed events using the `plotCylindersCI` function. This can help us identify cylinders that contain a significantly higher number of events than what would be expected based on the baseline.

```
plotCylindersCI(cylinders )
```

We then apply the function `warning.score2` to all grid locations to extract warning scores given the cylinders.
```
tmp = apply(d1, 1, FUN=warning.score2, 20, cylinders)
d1[, 'retrospective'] = tmp[1,]
head(d1)
```


## Prospective usage 1

`rancover` can be also prospectively, the same as what you would have got if you'd done it in realtime.
Draw cylinders up until a partial date  `TT`  with `only.last=TRUE` option and  apply `warning.score2` at each partial date. Setting `only.last=TRUE` generate cylinders that do not extend past the time `TT`.
These cyliders cover present or past events and extend up until the current day (and not in the future).


```
for (TT in tt){
  cylinders = CreateCylinders(observation.matrix,
                            baseline.matrix, range(tt), n.cylinders = 1000,
                            coord.df = d1, size_factor = 4, GT = 5, only.last = T, TT = TT)
  colname =  paste0('warning.score.', as.character(TT))
  tmp = apply(d1, 1, FUN=warning.score2, TT, cylinders)
  d1[,colname] = tmp[1,]
}
head(d1)
```
 
## Vignette Info


```{r}
sessionInfo()
```


## References

<a id="1">[1]</a>
M. Cavallaro, J. Coelho, D. Ready, V. Decraene, T. Lamagni, N. D. McCarthy, D. Todkill, M. J. Keeling (2022)
Cluster detection with random neighbourhood covering: Application to invasive Group A Streptococcal disease. PLoS Comput Biol 18(11): e1010726.
https://doi.org/10.1371/journal.pcbi.1010726



<a id="1">[2]</a>
M. Cavallaro, L. Dyson, M. J. Keeling (2023) Manuscript in preparation.
