---
title: "rancovr quickstart 1"
author: "Massimo Cavallaro"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style



## Create dataset

```{r}
x <- seq(0, 10, length.out = 20)
y <- seq(0, 10, length.out = 20)
tt = seq(1, 20)
d1 <- expand.grid(x = x, y = y)
d1$names = paste('loc', as.character(1:20), as.character(1:10), sep = '-')
baseline.matrix = sapply(tt, function( . ){
  exp(- (d1$x - 5)^2 - (d1$y - 5)^2 )  *  cos( 0.2 * . )^2} )
rownames(baseline.matrix) = d1$names
colnames(baseline.matrix) = as.character(tt)
baseline.matrix[1:5, 1:5]
```


##

```{r, fig.show='hold'}
image(x, y, matrix(baseline.matrix[,1], ncol=20))
image(x, y, matrix(baseline.matrix[,2], ncol=20))
```

##

```{r}
observation.matrix = matrix(
  rpois(length(baseline.matrix), baseline.matrix),
  ncol = ncol(baseline.matrix))

rownames(observation.matrix) = d1$names
colnames(observation.matrix) = as.character(tt)

observation.matrix[30:35, 10:15] = observation.matrix[30:35, 10:15] + rpois(36, 1)
```

```{r}
plot(colSums(observation.matrix))
lines(colSums(baseline.matrix))
```



```{r, fig.show='hold'}
t = 9
image(x, y, matrix(observation.matrix[,t], ncol=20), ncol=20
      , breaks = 1:5 - 0.5, main = paste0('t=', as.character(t)))
t = 15
image(x, y, matrix(observation.matrix[,t], ncol=20), main = paste0('t=', as.character(t)))
```


Retrospective usage. Draw all cylinders at once up until a date `TT`    with `only.last=FALSE` option and apply `warning.score2`:

```{r}
cylinders = CreateCylinders(observation.matrix,
                            baseline.matrix, range(tt), n.cylinders = 1000,
                            coord.df = d1, size_factor = 4, GT = 5)
head(cylinders)
plotCylindersCI(cylinders )
```


```{r}
tmp = apply(d1, 1, FUN=warning.score2, 20, cylinders)
d1[, 'retrospective'] = tmp[1,]
head(d1)
```


Prospective usage. Draw cylinders up until a partial date  `TT`  with `only.last=TRUE` option and  apply `warning.score2` at each partial date:

```{r}
for (TT in tt){
  cylinders = CreateCylinders(observation.matrix,
                            baseline.matrix, range(tt), n.cylinders = 1000,
                            coord.df = d1, size_factor = 4, GT = 5, only.last = T, TT = TT)
  colname =  paste0('warning.score.', as.character(TT))
  tmp = apply(d1, 1, FUN=warning.score2, TT, cylinders)
  d1[,colname] = tmp[1,]
}
head(d1)
```



## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
